\documentclass[a5paper]{article}
\input{../tex-preamble/amsthm.tex}
\input{../tex-preamble/anki.tex}
\input{../tex-preamble/general.tex}
\input{../tex-preamble/math.tex}
\input{../tex-preamble/problems.tex}
\input{../tex-preamble/colors/tao-yang.tex}
\geometry{
  lmargin=1.5cm,
  rmargin=1.5cm,
  tmargin=2cm,
  bmargin=2cm
}

\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{arrows,decorations.markings}

\usepackage{stmaryrd}

\usepackage{titling}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[L]{\theauthor}
\fancyhead[C]{\thetitle}
\fancyhead[R]{\thedate}

\newcommand{\wt}{\widetilde}

\begin{document}
\title{Awodey: Duality}
\author{Langston Barrett}
\date{Fall 2017}

\problemnum{2}
\begin{theorem*}
	The free monoid functor preserves binary coproducts.
  Specifically, for any sets $A,B\in\Obj(\Set)$, if $\caF:\Set\to\Mon$ denotes
  the free monoid functor, there is an isomorphism
  \begin{equation*}
    \caF(A)+\caF(B)\cong \caF(A+B)
  \end{equation*}
\end{theorem*}
\begin{proof}
  Let $f_A,f_B$ be given and label the inclusions into coproducts $i_A,i_B$ and
  the inclusions of generators $j_A,j_B,j_{A+B}$ as in the following diagram:
  \begin{center}
    \begin{tikzcd}[sep=large, row sep=large, column sep=large] 
      A \arrow[r, tail, "j_A"]
        \arrow[dr, tail, "i_A"]
        &
        \caF(A) \arrow[drr, Rightarrow, bend left=15, "f_A"]
        & {} & {}\\
      {} & A+B \arrow[r, tail, "j_{A+B}"] & {} \caF(A+B) & N\\
      B \arrow[r, tail, "j_B"]
        \arrow[ur, tail, "i_B"]
        & \caF(B) \arrow[urr, Rightarrow, bend right=15, "f_B"]
        & {} & {}
    \end{tikzcd}
  \end{center}
  Where double arrows denote arrows in $\Mon$, rather than just in $\Set$.
  The UMP of the free monoid guarantees monoid homomorphisms
  \begin{align*}
    k_A:\caF(A)\rightarrowtail \caF(A+B) &&\text{induced by}&&j_{A+B}\circ i_A \\
    k_B:\caF(B)\rightarrowtail \caF(A+B) &&\text{induced by}&&j_{A+B}\circ i_B
  \end{align*}
  such that
  \begin{align}
    |k_A| \circ j_A &= j_{A+B}\circ i_A \label{eq:ka} \\
    |k_B| \circ j_B &= j_{A+B}\circ i_B \label{eq:kb}
  \end{align}
  uniquely. They are monos because they are equal as set maps to the composition
  of two monos. The maps $|f_A|\circ j_A:A\to N$ and $|f_B|\circ j_B:B\to N$
  induce a set map $g:A+B\to N$ such that
  \begin{align}
    g\circ i_A &= |f_A|\circ j_A \label{eq:ga} \\
    g\circ i_B &= |f_B|\circ j_B \label{eq:gb}
  \end{align}
  So far, we have that the following diagram commutes:
  \begin{center}
    \begin{tikzcd}[sep=large, row sep=large, column sep=large] 
      A \arrow[r, tail, "j_A"]
        \arrow[dr, tail, "i_A"]
        &
        \caF(A) \arrow[drr, Rightarrow, bend left=15, "f_A"]
                \arrow[dr, dashed, Rightarrow, "k_A"]
        & {} & {}\\
        {} & A+B \arrow[r, tail, "j_{A+B}"] \arrow[drru, dashed, bend right=20, "g"]
        & \caF(A+B) % \arrow[r,dashed,Rightarrow, "h"]
        & N\\
      B \arrow[r, tail, "j_B"]
        \arrow[ur, tail, "i_B"]
        & \caF(B) \arrow[urr, Rightarrow, bend right=15, "f_B"]
                  \arrow[ur, dashed, Rightarrow, "k_B"]
        & {} & {}
    \end{tikzcd}
  \end{center}
  The UMP of the free monoid produces a monoid homomorphism $h:\caF(A+B)\to N$
  such that
  \begin{equation}\label{eq:umph}
    |h|\circ j_{A+B}=g
  \end{equation}
  It remains to show that
  \begin{align*}
    h\circ k_A = f_A &&\text{and}&& h\circ k_B = f_B
  \end{align*}
  We have
  \begin{align*}
    |h|\circ |k_A|\circ j_A
    &= |h|\circ j_{A+B}\circ i_A &&\quad\text{Equation \eqref{eq:ka}}\\
    &= g\circ i_A &&\quad\text{Equation \eqref{eq:umph}} \\
    &= |f_A|\circ j_A &&\quad\text{Equation \eqref{eq:ga}}
  \end{align*}
  and
  \begin{align*}
    |h|\circ |k_B|\circ j_B
    &= |h|\circ j_{A+B}\circ i_B &&\quad\text{Equation \eqref{eq:kb}}\\
    &= g\circ i_B &&\quad\text{Equation \eqref{eq:umph}} \\
    &= |f_B|\circ j_B &&\quad\text{Equation \eqref{eq:gb}}
  \end{align*}
  Which, combined with the uniqueness conditions of the respective UMPs, gives
  the required equality.
\end{proof}

\problemnum{6}

\begin{theorem*}
	The category of monoids ($\Mon$) has all equalizers.
\end{theorem*}
\begin{proof}
	(Equalizers) Let $A,B\in\Obj(\Mon)$ and
  \begin{tikzcd}[cramped,sep=small] f_1,f_2:A
    \arrow[r,yshift=2pt]\arrow[r,yshift=-2pt] & B \end{tikzcd}. We need to
  give a monoid $E$ together with an arrow $e:E\rightarrowtail A$ satisfying
  the UMP. As in $\Set$, the set $E$ is the subset of $A$ where $f_1$ and $f_2$
  agree, and $e$ is the inclusion.
  \begin{equation*}
    E\coloneqq\braces{a\in A|f_1(a)=f_2(a)}
  \end{equation*}
  By definition, $f_1\circ e=f_2\circ e$. We need to show that this is in fact
  an object of $\Mon$, i.e.\ a submonoid of $A$. Let $x,y\in E$, meaning
  $f_1(x)=f_2(x)$ and $f_1(y)=f_2(y)$.
  \begin{align*}
    f_1(xy)
    &= f_1(x)f_1(y)
    &&\quad f_1\text{ is a homomorphism} \\
    &= f_2(x)f_2(y)
    &&\quad \text{Assumption on }x,y \\
    &= f_2(xy)
    &&\quad f_2\text{ is a homomorphism}
  \end{align*}
  Additionally, $E$ is non-empty, since by the definition of a monoid morphism,
  $f_1(u_A)=f_2(u_A)=u_B$ where $u_A\in A,u_B\in B$ are the identities. It
  remains to show that any other map that equalizes $f_1$ and $f_2$ factors
  through $e$. Let $z:Z\to A$ such that $f_1\circ z=f_2\circ z$ and let
  $x\in Z$. Then $f_1(z(x))=f_2(z(x))$, so $z(x)\in E$ (more specifically, there
  is some $x'\in E$ such that $e(x')=z(x)$). Thus, $Z$ is a submonoid of $E$ and
  $z$ factors through $e$ via the inclusion $Z\rightarrowtail E$.
\end{proof}

\problemnum{7}

\begin{theorem*}
	The coproduct of projective objects is projective. 
\end{theorem*}
\begin{proof}
  Let $\bfC$ be a category with binary coproducts and let $P_1,P_2\in\Obj(\bfC)$
  be projective. Let $P_1+P_2$ denote their coproduct, and assume as given
  $f:P_1+P_2\to X$ and $e:E\twoheadrightarrow X$. By the UMP of the coproduct,
  we have
  \begin{align*}
    f_1:P_1&\longrightarrow X  &&\text{such that}&& f\circ i_1 = f_1 \\
    f_2:P_2&\longrightarrow X  &&\text{such that}&& f\circ i_2 = f_2 
  \end{align*}
  and since $P_1$ and $P_2$ are projective, we get
  \begin{align*}
    \ov{f_1}:P_1\longrightarrow E &&\text{such that}&& e\circ \ov{f_1} = f_1 \\
    \ov{f_2}:P_2\longrightarrow E &&\text{such that}&& e\circ \ov{f_2} = f_2
  \end{align*}
  which, by the UMP of the coproduct, yield
  \begin{align*}
    \ov{f}:P_1+P_2\longrightarrow E
  \end{align*}
  such that $\ov{f}\circ i_1=\ov{f_1}$ and $\ov{f}\circ i_2=\ov{f_2}$. All taken
  together, the following diagram commutes:
  \begin{center}
    \begin{tikzcd}[row sep=large]
      {} & X & {}\\
      {} & E \arrow[u, twoheadrightarrow, "e"]& \\
      P_1\arrow[r, "i_1", swap] \arrow[ur,dashed, "\ov{f_1}"]
         \arrow[uur, dashed, "f_1", bend left=15] %
      & P_1+P_2 \arrow[u, "\ov{f}"] &
      P_2\arrow[l, "i_2"]\arrow[ul,dashed, "\ov{f_2}", swap]
         \arrow[uul, dashed, "f_2", swap, bend right=15]
    \end{tikzcd}
  \end{center}
  Reading off the diagram,
  \begin{align*}
    \ov{f} \circ i_1 = \ov{f_1}
    &\implies (e\circ\ov{f})\circ i_1 = e\circ\ov{f_1}
    &&\quad\text{Left composition with }e \\
    &\implies (e\circ\ov{f})\circ i_1 = f_1
    &&\quad\text{Mapping property of }\ov{f_1} \\
    \ov{f} \circ i_2 = \ov{f_2}
    &\implies (e\circ\ov{f})\circ i_2 = e\circ\ov{f_2}
    &&\quad\text{Left composition with }e \\
    &\implies (e\circ\ov{f})\circ i_2 = f_2
    &&\quad\text{Mapping property of }\ov{f_2}
  \end{align*}
  However, by the UMP of the coproduct, $f$ is the unique map making the above
  equations commute.
  Thus, $e\circ\ov{f}=f$, which is exactly the mapping property of a
  projective.
\end{proof}

\end{document}