\documentclass[a5paper]{article}
\input{./exercise-header.tex}
\begin{document}
\title{Hatcher: Covering Spaces}
\author{Langston Barrett}
\date{Fall 2017}

\begin{definition*}
	A \textbf{covering space} of a topological space $(X,\tau)$ consists of the
  following data:
  \begin{itemize}
    \itemsep0em
    \item A space $(\widetilde{X},\widetilde{\tau})$,
    \item a continuous map $p:\widetilde{X}\to X$,
    \item and an open cover $\braces{U_\alpha}_{\alpha\in A}$ of $X$
  \end{itemize}
  subject to the condition that for all $\alpha\in A$, the preimage
  $p^{-1}(U_\alpha)$ is a disjoint union of open sets in $\widetilde{X}$, each of
  which is mapped homeomorphically onto $U_\alpha$ by $p$.
\end{definition*}
\problemnum{1}
\begin{proposition*}
	Let $\wt{X}$ be a covering space of $X$ with map $p:\widetilde{X}\to X$ and
  open cover $\braces{U_\alpha}_{\alpha\in A}$. If $A$ is a subspace of $X$, then
  $p^{-1}(A)$ is a covering space of $A$ with map $q:p^{-1}(A)\to A$
  the restriction of $p$ and cover
  $\braces{U_\beta}\coloneqq \braces{A\cap U_\alpha}_{\alpha\in A}$.
\end{proposition*}
\begin{proof}
  Since $\braces{U_\alpha}$ covers $X$, it certainly covers $A$. Since the
  intersection of an open set with $A$ is open in the subspace topology,
  $\braces{U_\beta}$ is an open cover of $A$.

  Consider the preimage $q^{-1}(U_\beta)$ for some $\beta$.
  Since the restriction of continuous maps to a subspace is continuous,
  $q^{-1}$ is continuous. Thus, $q^{-1}(U_\beta)$ is the disjoint union of open
  sets in $p^{-1}(A)$ and each component is mapped homeomorphically onto
  $U_\beta$ by $q$. We have given the required data and shown that it follows
  the conditions, so $p^{-1}(A)$ is a covering space of $A$.
\end{proof}

\problemnum{2}

\begin{lemma*}
  The following is a basis for a disjoint union $\coprod_i B_i$:
  \begin{equation*}
    \caB \coloneqq \braces{\iota_j(U)|U\text{ is open in }B_j}
  \end{equation*}
  where $\iota_j:B_j\hookrightarrow \coprod_i B_i$ is the canonical inclusion.
\end{lemma*}
\begin{proof}
	This is a restatement of the universal property of the coproduct.
\end{proof}

\begin{lemma*}
  In $\Top$, finite products commute with arbitrary coproducts. Namely for all
  spaces $A$ and a disjoint union $\coprod_i B_i$,
  \begin{equation*}
    A\times \coprod_i B_i\cong \coprod_i (A\times B_i)
  \end{equation*}
\end{lemma*}
\begin{proof}
  Recall that the coproduct is an indexed union, a set of pairs where the first
  element of the pair is a ``tag'' indicating which set that item originated from.
  \begin{align*}
    \phi:A\times\coprod_i B_i &\longrightarrow \coprod_i (A\times B_i) \\
    (a,(i,b))&\longmapsto (i,(a,b)) \\
  \intertext{with inverse}
    A\times\coprod_i B_i &\longleftarrow \coprod_i (A\times B_i):\phi^{-1} \\
    (a,(i,b))&\longmapsfrom (i,(a,b)) 
  \end{align*}
  To prove continuity of a map into a space, it suffices to prove that the
  inverse images of basis elements are open. As we have bases for both the
  product and coproduct topologies, it suffices to show that 
  $\phi^{-1}(j,U\times V)$ is open where $U$ is open in $A$ and $V$ is open in
  $B_j$. From the definition of $\phi$ and $\phi^{-1}$, we obtain
  \begin{equation*}
    \phi^{-1}(j,U\times V) = (U,(j,V))
  \end{equation*}
  which is open by definition of the product topology. A similar argument shows
  that $\phi^{-1}$ is continuous.
\end{proof}

\begin{proposition*}
	If $\tilde{X}$, $p:\tilde{X}\to X$, $\braces{U_\alpha}$ and $\tilde{Y}$,
  $q:\tilde{Y}\to Y$, $\braces{V_\beta}$ are covering spaces, then the product
  $\tilde{X}\times\tilde{Y}$ covers $X\times Y$.
\end{proposition*}
\begin{proof}
  If $(x,y)\in X\times Y$, then there exist $U_\alpha\ni x$ and $V_\beta\ni y$
  in the open covers of $X$ and $Y$. Their product $U_\alpha\times V_\beta$
  covers $(x,y)$ and is open in the product topology. Therefore, the pairwise
  cartesian product of the open covers on $X$ and $Y$ defines an open cover on
  $X\times Y$.

  Let $\langle p,q \rangle:\widetilde{X}\times \widetilde{Y}\to X\times Y$ be
  the product of maps $p$ and $q$, and consider the preimage
  \begin{align*}
    \langle p,q \rangle^{-1}(U_\alpha\times V_\beta)
    &= \braces{(x,y)\in \wt{X}\times\wt{Y}\Big|\langle p,q \rangle(x,y)\in U_\alpha\times V_\beta} \\
    &= \braces{(x,y)\in \wt{X}\times\wt{Y}\Big|(p(x),q(y))\in U_\alpha\times V_\beta} \\
    &= \braces{(x,y)\in \wt{X}\times\wt{Y}\Big|p(x)\in U_\alpha\text{ and }q(y)\in V_\beta} \\
    &= p^{-1}(U_\alpha)\times q^{-1}(V_\beta)
  \end{align*}
  By continuity of $p$ and $q$, this is an open set in the product topology on
  $\wt{X}\times \wt{Y}$.
  It remains to show that it is the disjoint union of sets
  mapped homeomorphically onto $U_\alpha\times V_\beta$ by $\langle p,q \rangle$.
  A slight generalization of the above lemma demonstrates that the inverse image
  is indeed a disjoint union of open sets. The definition of
  $\langle p,q \rangle$ ensures that the mapping is homeomorphic.
\end{proof}

\problemnum{3}

% \begin{lemma*}
% 	If $\wt{X}$ covers $X$ via $p:\wt{X}\to X$, then $p$ is an open map.
% \end{lemma*}
% \begin{proof}
% 	Let $\wt{U}\subseteq\wt{X}$ be open. By the definition of a covering space,
%   there is a distinguished open cover $\braces{U_\alpha}$ of $X$. Then
%   $p(\wt{U})$ is the subset of a union of some of these covering sets.
%   Additionally, the intersection
% \end{proof}

\begin{proposition*}
	Let $\wt{X}$ be a covering space with map $\wt{X}\to X$ and suppose
  $p^{-1}(x)$ is finite and nonempty for all $x\in X$. Then $X$ is compact
  Hausdorff if and only if $\wt{X}$ is.
\end{proposition*}
\begin{proof}
	($\Rightarrow$ Hausdorff) Assume $X$ is Hausdorff. Let $\wt{x},\wt{y}$ be
  distinct points in $\wt{X}$ and let $x\coloneqq p(\wt{x})$
  and $y\coloneqq p(\wt{y})$. Since $X$ is Hausdorff, there exist non-overlapping
  neighborhoods $U$ of $x$ and $V$ of $y$. Call the intersections of these
  neighborhoods with the open cover guaranteed by the covering space $U'$ and
  $V'$, respectively. Then $p^{-1}(U')$ and $p^{-1}(V')$ are disjoint unions of
  open and distinct sets in $\wt{X}$, and contain $\wt{x}$ and $\wt{y}$
  respectively.

	($\Leftarrow$ Hausdorff) Assume $\wt{X}$ is Hausdorff. Let $x,y$ be
  distinct points in $X$. By the nonemptiness assumption, there exist
  distinct $\wt{x},\wt{y}\in\wt{X}$ such that $p(\wt{x})=x$ and $p(\wt{y})=y$.
  By the axioms of covering spaces, $p^{-1}(x)$ and $p^{-1}(y)$ are the disjoint
  unions of open sets. Let $\wt{U}_x,\wt{U}_y$ be the open sets containing
  $\wt{x}$ and $\wt{y}$, respectively. Since $\wt{X}$ is Hausdorff, there exist
  non-overlapping open sets $\wt{U}_x'$ and $\wt{U}_y'$ contained within
  $\wt{U}_x$ and $\wt{U}_y$, respectively. Since homeomorphisms are open maps,
  $p(\wt{U}_x')$ and $q(\wt{U}_y')$ are open in $X$ and contain $x$ and $y$,
  respectively. 
\end{proof}

\end{document}